Robust Quantum Optimal Control with Trajectory Optimization

### 2021-03-13 paper/5static.tex
dave comments up to sec iv and readme updates

Whereas the sampling method penalizes the deviations of the sample states
from the target state, the unscented sampling method
penalizes the deviations of the sample states from the nominal state
\cite{howell2020direct, lee2013sigma,
  thangavel2020robust}. Accordingly, the cost function we add
to the objective \eqref{eq:costfun} takes the form
$\sum_{k, j} c_{k} (\psi^{j}_{k} - \psi_{k})^{T}
(\psi^{j}_{k} - \psi_{k})$, where $c_{k}$ is a
constant we supply, $\psi_{k}$ is
the evolved initial state (nominal state), and $\psi^{j}_{k}$ is a sample state
that evolves under a modified Hamiltonian similar to that in the sampling method.
We omit bra-ket notation here to emphasize that
the states are real vectors and are given by the right-hand-side of the
complex-to-real isomorphism \eqref{eq:isomorphism}.
The sample states are chosen to encode a unimodal distribution over
the $2n$ elements of the nominal state, modeling the uncertainty in the state
as a result of the uncertainty in the parameter. We use the unscented transform
\cite{julier2004unscented, uhlmann1995dynamic}
to accurately propagate the mean and covariance of this distribution between
knot points, or equivalently, through the transformation of the TDSE \eqref{eq:tdse}.
Unlike the sampling method, the cost function for the unscented sampling method
is sensitive to global phases. Accordingly, we
do not observe a performance increase when
using more than one initial state. 
A detailed procedure for the unscented transformation is given
in Appendix \ref{appendix:unscented}.

The derivative method penalizes the sensitivity of the state
to the uncertain parameter, which is encoded in the $l$\textsuperscript{th}-order
state derivative $\ket*{\partial_{f_{q}}^{l} \psi} \equiv \partial_{f_{q}}^{l} \ket*{\psi}$.
In the $m$\textsuperscript{th}-order
derivative method, we append all state derivatives of order $1, \dots, m$
to the augmented state vector \eqref{eq:astatecontrols}
for each initial state.
We obtain the state derivatives at each knot point by performing forward-mode
differentiation on the TDSE \eqref{eq:tdse}.
For example, the dynamics for the $1$\textsuperscript{st}-order derivative method are:
\begin{align}
  i \hbar \frac{d}{dt} \ket{\psi} &= H \ket{\psi},\\
  i \hbar \frac{d}{dt} \ket{\partial_{f_{q}}\psi} &=
  H \ket{\partial_{f_{q}} \psi} +
  (\partial_{f_{q}} H) \ket{\psi}.
  \label{eq:d1dyn}
\end{align}
We integrate the coupled ODEs with exponential
integrators in the discrete dynamics function \eqref{eq:dyn_con},
see Appendix \ref{appendix:derivative}.
While the state $\ket*{\psi}$ has unit norm,
the state derivatives $\ket*{\partial^{m}_{f_{q}} \psi}$ need not, as is evident
from the non-unitary dynamics \eqref{eq:d1dyn}.
We penalize the norms of the state derivatives
in the objective \eqref{eq:costfun} by setting the corresponding elements
of the target augmented state to zero. Intuitively, this corresponds to penalizing
the sensitivity of each state element to the uncertain parameter. As was the case for
the unscented sampling
method, we do not observe a performance increase when using more than one initial state
for the derivative method.
For runtimes of the three robust control techniques,
consult Appendix \ref{appendix:time}.

@@

The analytic gate corresponds to
idling at the flux frustration point $a_{k} = 0 \ \forall \ k$, see Figure
\ref{fig:statica}. Its gate time $1 / 4 f_{q} \sim 18\textrm{ns}$
is the shortest possible for a $Z/2$ gate on the device.
The gate's erroneous rotation angle
$2 \pi \delta f_{q} / 4 f_{q}$ is linear in the
qubit frequency detuning, resulting in a gate error that is quadratic
in the detuning.
At a one-percent detuning ($\lvert \delta f_{q} / f_{q} \rvert = 1\%$),
the gate error is $\sim 4.7 \cdot 10^{-5}$,
which is sufficient for quantum error correction.

@@

@@

The two derivative methods converge on qualitatively similar flux pulses that
idle near the flux frustration point and use fast triangle movements at the boundaries,
similar to the flux pulse produced by the unscented sampling method.
For both methods, the gate error at a one-percent qubit frequency detuning
decreases super-linearly in the gate time.
For the 1\textsuperscript{st}-order method, the gate error at a one-percent detuning 
reaches $10^{-7}$ at the Larmor period $1 / f_{q} \sim 72$ns,
see Figure \ref{fig:staticc}.
This result mimics the
ability of composite pulses to mitigate parameter uncertainty errors to arbitrary
order with sufficiently many pulses \cite{merrill2014progress}.
It is difficult to choose an appropriate composite pulse
for the problem studied here due to our Hamiltonian and experimental constraints.
We propose comparisons between composite pulses and numerical techniques
for future work.

>>>>

Whereas the sampling method penalizes the deviations of the sample states
from the target state, the unscented sampling method
penalizes the deviations of the sample states from the nominal state
\cite{howell2020direct, lee2013sigma,
  thangavel2020robust}. Accordingly, the cost function we add
to the objective \eqref{eq:costfun} takes the form
$\sum_{k, j} c_{k} (\psi^{j}_{k} - \psi_{k})^{T}
(\psi^{j}_{k} - \psi_{k})$, where $c_{k}$ is a
constant we supply, $\psi_{k}$ is
the evolved initial state (nominal state), and $\psi^{j}_{k}$ is a sample state
that evolves under a modified Hamiltonian similar to that in the sampling method.
We omit bra-ket notation here to emphasize that
the states are real vectors and are given by the right-hand-side of the
complex-to-real isomorphism \eqref{eq:isomorphism}.
The sample states are chosen to encode a unimodal distribution over
the $2n$ elements of the nominal state, modeling the uncertainty in the state
as a result of the uncertainty in the parameter. We use the unscented transform
\cite{julier2004unscented, uhlmann1995dynamic}
to accurately propagate the mean and covariance of this distribution between
time steps, or equivalently, through the transformation of the TDSE \eqref{eq:tdse}.
Unlike the sampling method, the cost function for the unscented sampling method
is sensitive to global phases. Accordingly, we
do not observe a performance increase when
using more than one initial state. 
A detailed procedure for the unscented transformation is given
in Appendix \ref{appendix:unscented}.

The derivative method penalizes the sensitivity of the state
to the uncertain parameter, which is encoded in the $l$\textsuperscript{th}-order
state derivative $\ket*{\partial_{f_{q}}^{l} \psi} \equiv \partial_{f_{q}}^{l} \ket*{\psi}$.
In the $m$\textsuperscript{th}-order
derivative method, we append all state derivatives of order $1, \dots, m$
to the augmented state vector \eqref{eq:astatecontrols}
for each initial state.
We obtain the state derivatives at each time step by performing forward-mode
differentiation on the TDSE \eqref{eq:tdse}.
For example, the dynamics for the $1$\textsuperscript{st}-order derivative method are:
\begin{align}
  i \hbar \frac{d}{dt} \ket{\psi} &= H \ket{\psi},\\
  i \hbar \frac{d}{dt} \ket{\partial_{f_{q}}\psi} &=
  H \ket{\partial_{f_{q}} \psi} +
  (\partial_{f_{q}} H) \ket{\psi}.
  \label{eq:d1dyn}
\end{align}
We integrate the coupled ODEs with exponential
integrators in the discrete dynamics function \eqref{eq:dyn_con},
see Appendix \ref{appendix:derivative}.
While the state $\ket*{\psi}$ has unit norm,
the state derivatives $\ket*{\partial^{l}_{f_{q}} \psi}$ need not, as is evident
from the non-unitary dynamics \eqref{eq:d1dyn}.
We penalize the norms of the state derivatives
in the objective \eqref{eq:costfun} by setting the corresponding elements
of the target augmented state to zero. Intuitively, this corresponds to penalizing
the sensitivity of each state element to the uncertain parameter. As was the case for
the unscented sampling
method, we do not observe a performance increase when using more than one initial state
for the derivative method.
For runtimes of the three robust control techniques,
consult Appendix \ref{appendix:time}.

@@

@@

The analytic gate corresponds to
idling at the flux frustration point $a_{k} = 0 \ \forall \ k$, see Figure
\ref{fig:statica}. Its gate time $1 / 4 f_{q} \sim 18\textrm{ns}$
is the shortest possible for a $Z/2$ gate on the device.
The gate's erroneous rotation angle
$2 \pi \delta f_{q} / 4 f_{q}$ is linear in the
qubit frequency detuning, resulting in a gate error that is quadratic
in the detuning.
At a one-percent detuning $\lvert \delta f_{q} / f_{q} \rvert = 1\%$,
the gate error is $\sim 4.7 \cdot 10^{-5}$,
which is sufficient for quantum error correction.

@@

The two derivative methods converge on qualitatively similar flux pulses that
idle near the flux frustration point and use fast triangle movements at the boundaries,
similar to the flux pulse produced by the unscented sampling method.
For both derivative methods, the gate error at a one-percent qubit frequency detuning
decreases super-linearly in the gate time.
For the 1\textsuperscript{st}-order method, the gate error at a one-percent detuning 
reaches $10^{-7}$ at the Larmor period $1 / f_{q} \sim 72$ns,
see Figure \ref{fig:staticc}.
This result mimics the
ability of composite pulses to mitigate parameter uncertainty errors to arbitrary
order with sufficiently many pulses \cite{merrill2014progress}.
It is difficult to choose an appropriate composite pulse
for the problem studied here due to our Hamiltonian and experimental constraints.
We propose comparisons between composite pulses and numerical techniques
for future work.

### 2021-03-13 paper/a1longitude.tex
dave comments up to sec iv and readme updates

We employ the Lindblad master equation
to compute the gate error due to depolarization.
This equation takes the form:
\begin{equation}
  \frac{d}{dt} \rho = \frac{-i}{\hbar} [H, \rho]
  + \sum_{i} \gamma_{i} (L_{i} \rho L_{i}^{\dagger}
  - \frac{1}{2} \{L_{i}^{\dagger} L_{i}, \rho\}),
\end{equation}
where $\rho = \ket{\psi}\bra{\psi}$ is the density matrix,
$[\cdot, \cdot]$ is the algebraic commutator, and $\{\cdot, \cdot \}$ is the algebraic
anti-commutator.
For depolarization, $\gamma_{\pm} = T_{\pm}^{-1}$,
$L_{\pm} = \sigma^{\pm} \equiv (\sigma_{x} \pm i \sigma_{y})/2$.
The depolarization times $T_{+} = T_{-} = 2 T_{1}$ are obtained at each knot point
from the spline shown in Figure \ref{fig:longitudeb}.
We obtain the $T_{1}$ values in this spline
by driving the qubit at the desired flux bias
and monitoring the resultant decay. For more details
on these measurements, consult \cite{zhang2020universal}.
Because $T_{1}$ depends on the flux $a(t)$, so do
the decay rates $\gamma_{i}$.
Integrating the master equation with time-dependent decay rates
provides a heuristic for how the gates we study may perform in
the experiment. 
This procedure may not be strictly correct
when the decay rates change much faster
than the relaxation time $\Delta t_{k} \ll T_{1}$,
which is the regime we are operating in. Standard
derivations of the Lindblad master equation do not account for
time-dependent decay rates \cite{manzano2020a}. A more thorough
treatment of this regime in future work would unlock new insights for
quantum computing platforms where decoherence is strongly
dependent on the control parameters.

So that we may use exponential integrators, we employ
the Vectorization/Choi-Jamiolkowski isomorphism \cite{Landi2018},
\begin{equation}
  \frac{d}{dt} \textrm{vec}({\rho}) = \hat{\mathcal{L}} \textrm{vec}({\rho}),
\end{equation}
\begin{equation}
  \begin{aligned}
    \hat{\mathcal{L}} &= -i(I \otimes H - H^{T} \otimes I)\\
    &+ \sum_{i} \gamma_{i}
    (L_{i}^{*} \otimes L_{i} - \frac{1}{2} (I \otimes L_{i}^{\dagger}L_{i}
    - L_{i}^{T}L_{i}^{*} \otimes I)),
  \end{aligned}
\end{equation}
where $\rho = \sum_{i, j} \alpha_{ij} \ket{i}\bra{j}$
and $\textrm{vec}(\rho) = \sum_{i, j} \alpha_{ij} \ket{i} \otimes \ket{j}$.
We use zero-order hold on the controls; equivalently, $a(t)$ is constant
on each integration interval $[t_{k}, t_{k + 1}]$. Therefore,
$H(a(t))$ and $\gamma_{i}(a(t))$ are also piecewise-constant.
So, the exact solution is
$\textrm{vec}(\rho_{k + 1}) = {\exp}{\textstyle(}\Delta t_{k}
\hat{\mathcal{L}}_{k}{\textstyle)} \textrm{vec}(\rho_{k})$.
This isomorphism transforms $(n \times n) \times (n \times n)$
matrix-matrix multiplications to $(n^{2} \times n^{2}) \times n^{2}$ matrix-vector
multiplications. For small $n$ and zero-order hold on the controls, we find that it is
faster to use an exponential integrator on the vectorized equation than to perform
Runge-Kutta on the unvectorized equation. The latter requires decreasing the integration
time step to maintain accuracy, resulting in more knot points.

>>>>

We employ the Lindblad master equation
to compute the gate error due to depolarization.
This equation takes the form:
\begin{equation}
  \frac{d}{dt} \rho = \frac{-i}{\hbar} [H, \rho]
  + \sum_{i} \gamma_{i} (L_{i} \rho L_{i}^{\dagger}
  - \frac{1}{2} \{L_{i}^{\dagger} L_{i}, \rho\}),
\end{equation}
where $\rho = \ket{\psi}\bra{\psi}$ is the density matrix,
$[\cdot, \cdot]$ is the algebraic commutator, and $\{\cdot, \cdot \}$ is the algebraic
anti-commutator.
For depolarization, $\gamma_{\pm} = T_{\pm}^{-1}$,
$L_{\pm} = \sigma^{\pm} \equiv (\sigma_{x} \pm i \sigma_{y})/2$.
The depolarization times $T_{+} = T_{-} = 2 T_{1}$ are obtained at each time step
from the spline shown in Figure \ref{fig:longitudeb}.
We obtain the $T_{1}$ values in this spline
by driving the qubit at the desired flux bias
and monitoring the resultant decay. For more details
on these measurements, consult \cite{zhang2020universal}.
Because $T_{1}$ depends on the flux $a(t)$, so do
the decay rates $\gamma_{\pm}$.
Integrating the master equation with time-dependent decay rates
provides a heuristic for how the gates we study may perform in
the experiment. 
This procedure may not be strictly correct
when the decay rates change much faster
than the relaxation time $\Delta t_{k} \ll T_{1}$,
which is the regime we are operating in. Standard
derivations of the Lindblad master equation do not account for
time-dependent decay rates \cite{manzano2020a}. A more thorough
treatment of this regime in future work would unlock new insights for
quantum computing platforms where decoherence is strongly
dependent on the control parameters.

So that we may use exponential integrators, we employ
the Vectorization/Choi-Jamiolkowski isomorphism \cite{Landi2018},
\begin{equation}
  \frac{d}{dt} \textrm{vec}({\rho}) = \hat{\mathcal{L}} \textrm{vec}({\rho}),
\end{equation}
\begin{equation}
  \begin{aligned}
    \hat{\mathcal{L}} &= -i(I \otimes H - H^{T} \otimes I)\\
    &+ \sum_{i} \gamma_{i}
    (L_{i}^{*} \otimes L_{i} - \frac{1}{2} (I \otimes L_{i}^{\dagger}L_{i}
    - L_{i}^{T}L_{i}^{*} \otimes I)),
  \end{aligned}
\end{equation}
where $\rho = \sum_{i, j} \alpha_{ij} \ket{i}\bra{j}$
and $\textrm{vec}(\rho) = \sum_{i, j} \alpha_{ij} \ket{i} \otimes \ket{j}$.
Because $a(t)$ is constant between time steps $[t_{k}, t_{k + 1})$,
so are $H(a(t))$ and $\gamma_{\pm}(a(t))$.
Therefore, the exact solution is
$\textrm{vec}(\rho_{k + 1}) = {\exp}{\textstyle(}\Delta t_{k}
\hat{\mathcal{L}}_{k}{\textstyle)} \textrm{vec}(\rho_{k})$.
The vector isomorphism transforms $(n \times n) \times (n \times n)$
matrix-matrix multiplications to $(n^{2} \times n^{2}) \times n^{2}$ matrix-vector
multiplications. For small $n$ and zero-order hold on the controls, we find that it is
faster to use an exponential integrator on the vectorized equation than to perform
Runge-Kutta on the unvectorized equation. The latter requires decreasing the interval $\Delta t_{k}$
to maintain accuracy, resulting in more time steps.

### 2021-03-13 paper/a2unscented.tex
resolved danny edits
dave comments up to sec iv and readme updates

\section{Unscented Transformation}
\label{appendix:unscented}
In this section, we outline the full unscented sampling procedure.
In the unscented sampling method, we consider a state $\psi \in \mathbb{R}^{2n}$, 
an uncertain parameter $\lambda \in \mathbb{R}^{d}$, and discrete dynamics
$\psi_{k + 1} = f(\psi_{k}, \lambda_{k})$.
We omit bra-ket notation to emphasize that
the state is a real vector, and it is given by the right-hand-side of the
isomorphism \eqref{eq:isomorphism}.
The nominal initial state is given by $\bar{\psi}_{1}$ with an associated
covariance matrix $P_{1} \in \mathbb{S}_{++}^{2n}$
which describes the uncertainty in the initial state.
We use the notation $\mathbb{S}_{++}^{m}$ to denote the
set of real, symmetric, and positive-definite $m \times m$ matrices.
$P_{1}$ is typically non-zero even if the state preparation error is negligible.
The uncertain parameter has zero-mean
and its distribution is given by the covariance matrix
$L_{k} \in \mathbb{S}_{++}^{d}$ at knot point $k$. The zero-mean assumption
is convenient for deriving the update procedure. A non-zero mean can be encoded
in the discrete dynamics function $f(\psi_{k}, \lambda_{k})$.

The initial $4n + 2d$ sample states and initial $4n + 2d$
uncertain parameters are sampled from the initial distributions,
\begin{equation}\label{eq:uupdate}
  \begin{bmatrix} \psi_{1}^{j} \\ \lambda_{1}^{j} \end{bmatrix} =
  \begin{bmatrix} \bar{\psi}_{1} \\ 0\end{bmatrix}
    \pm \beta \sqrt{\begin{bmatrix} P_{1} & 0\\ 0 & L_{1}\end{bmatrix}}^{\; j}.
\end{equation}
$\beta$ is a hyperparameter that controls the spacing of the covariance contour.
The $(\pm)$ is understood to take $(+)$ for $j \in \{1, \dots, 2n + d\}$ and $(-)$ for
$j \in \{2n + d + 1, \dots, 4n + 2d\}$. We use the Cholesky factorization
to compute the square root of the
joint covariance matrix, though other methods
such as the principal square root may be employed.
The superscript on the matrix square root indicates the $j$\textsuperscript{th}
column (mod $2n + d$) of the lower triangular Cholesky factor.
Then, the sample states are normalized,
\begin{equation}\label{eq:unormalize}
  \psi_{1}^{j} \rightarrow \frac{\psi_{1}^{j}}{\sqrt{{\psi_{1}^{j}}^{T} \psi_{1}^{j}}}.
\end{equation}
The sample states are propagated to the next knot point,
\begin{equation}\label{eq:upropagate}
  \psi^{j}_{2} = f(\psi^{j}_{1}, \lambda^{j}_{1}).
\end{equation}
The mean and covariance of the sample states are computed,
\begin{align}
  \bar{\psi}_{2} &= \frac{1}{4n + 2d} \sum_{j = 1}^{4n + 2d} \psi_{2}^{j},\\
  P_{2} &= \frac{1}{2 \beta^{2}} \sum_{j = 1}^{4n + 2d}
  (\psi^{j}_{2} - \bar{\psi}_{2})(\psi^{j}_{2} - \bar{\psi}_{2})^{T}.
\end{align}
The sample states are then resampled and propagated to the next knot point using
\eqref{eq:uupdate}, \eqref{eq:unormalize}, and \eqref{eq:upropagate}. Our
choice of sample states (sigma points) follows equation 11 of
\cite{julier2004unscented}.
Prescriptions that require fewer sigma points exist \cite{julier2002reduced}.

>>>>

\section{Unscented Sampling Method}
\label{appendix:unscented}
In this section, we outline the full unscented sampling procedure.
We consider a state $\psi \in \mathbb{R}^{2n}$, 
an uncertain parameter $\lambda \in \mathbb{R}^{d}$, and discrete dynamics
$\psi_{k + 1} = f(\psi_{k}, \lambda_{k})$.
We omit bra-ket notation to emphasize that
the state is a real vector, and it is given by the right-hand-side of the
complex-to-real isomorphism \eqref{eq:isomorphism}.
The nominal initial state is given by $\bar{\psi}_{1}$ with an associated
covariance matrix $P_{1} \in \mathbb{S}_{++}^{2n}$
which describes the uncertainty in the initial state.
We use the notation $\mathbb{S}_{++}^{m}$ to denote the
set of real, symmetric, and positive-definite $m \times m$ matrices.
By the positive-definite requirement, $P_{1}$ must be non-zero even if the state-preparation error
is negligible.
The uncertain parameter has zero-mean
and its distribution is given by the covariance matrix
$L_{k} \in \mathbb{S}_{++}^{d}$ at time step $k$. The zero-mean assumption
is convenient for deriving the update procedure. A non-zero mean can be encoded
in the discrete dynamics function $f(\psi_{k}, \lambda_{k})$.

The initial $4n + 2d$ sample states and initial $4n + 2d$
uncertain parameters are sampled from the initial distributions,
\begin{equation}\label{eq:uupdate}
  \begin{bmatrix} \psi_{1}^{j} \\ \lambda_{1}^{j} \end{bmatrix} =
  \begin{bmatrix} \bar{\psi}_{1} \\ 0\end{bmatrix}
    \pm \beta \sqrt{\begin{bmatrix} P_{1} & 0\\ 0 & L_{1}\end{bmatrix}}^{\; j}.
\end{equation}
$\beta$ is a hyperparameter that controls the spacing of the covariance contour.
The $(\pm)$ is understood to take $(+)$ for $j \in \{1, \dots, 2n + d\}$ and $(-)$ for
$j \in \{2n + d + 1, \dots, 4n + 2d\}$. We use the Cholesky factorization
to compute the square root of the
joint covariance matrix, though other methods
such as the principal square root may be employed.
The superscript on the matrix square root indicates the $j$\textsuperscript{th}
column (mod $2n + d$) of the lower triangular Cholesky factor.
Then, the sample states are normalized,
\begin{equation}\label{eq:unormalize}
  \psi_{1}^{j} \rightarrow \frac{\psi_{1}^{j}}{\sqrt{{\psi_{1}^{j}}^{T} \psi_{1}^{j}}}.
\end{equation}
The sample states are propagated to the next time step,
\begin{equation}\label{eq:upropagate}
  \psi^{j}_{2} = f(\psi^{j}_{1}, \lambda^{j}_{1}).
\end{equation}
The mean and covariance of the sample states are computed,
\begin{align}
  \bar{\psi}_{2} &= \frac{1}{4n + 2d} \sum_{j = 1}^{4n + 2d} \psi_{2}^{j},\\
  P_{2} &= \frac{1}{2 \beta^{2}} \sum_{j = 1}^{4n + 2d}
  (\psi^{j}_{2} - \bar{\psi}_{2})(\psi^{j}_{2} - \bar{\psi}_{2})^{T}.
\end{align}
The sample states are then resampled and propagated to the next time step using
\eqref{eq:uupdate}, \eqref{eq:unormalize}, and \eqref{eq:upropagate}. Our
choice of sample states (sigma points) follows equation 11 of
\cite{julier2004unscented}.
Prescriptions that require fewer sigma points exist \cite{julier2002reduced}.
